Polyester RNA-seq simulation study

### 2018-09-18 datasets/RNA-Seq/polyester-simulation.Rmd
Standardize alpha scale in polyester
Remove mean covariate; add weak covariate to Polyester

# results files that will be generated below
resfile_n5 <- file.path(resdir, "polyester-results-null5.rds")
resfile_n10 <- file.path(resdir, "polyester-results-null10.rds")
resfile_d5 <- file.path(resdir, "polyester-results-de5.rds")
resfile_d10 <- file.path(resdir, "polyester-results-de10.rds")

resfile_n5_uninfCov <- file.path(resdir, "polyester-results-null5-uninfCov.rds")
resfile_n10_uninfCov <- file.path(resdir, "polyester-results-null10-uninfCov.rds")
resfile_d5_uninfCov <- file.path(resdir, "polyester-results-de5-uninfCov.rds")
resfile_d10_uninfCov <- file.path(resdir, "polyester-results-de10-uninfCov.rds")

@@

```{r}
#' @param sampleSize is the number of samples in each condition
#' @param nDE is the number of DE genes
#' @param bd is the bench design object
#' @param uninformativeCovariate logical indicating whether to replace overall mean 
#'  independent covariate with a randomly generated (uninformative) covariate. 
#'  Default is FALSE.
#' @param rseed random seed
#' @param ngenes the number of genes to simulate. If this number is greater than 
#'  the number of genes in the count matrix, then the polyester parameters will be
#'  drawn with replacement from the available pool in order to obtain the requested
#'  number of genes.
simulateOneSplit <- function(X, rseed, 
                             nDE, sampleSize, bd, uninformativeCovariate = FALSE,
                             ngenes = nrow(counts)){
  uninf_covar <- rnorm(ngenes)

@@

@@

  # filter genes with mostly zeroes
  pzero = rowSums(dat_alt==0)/ncol(dat_alt)
  dat_alt <- dat_alt[pzero < 0.5,]
  truth <- rep(FALSE, nrow(dat_alt))
  log2FC <- log2FC[pzero < 0.5]
  mu <- mu[pzero < 0.5]
  p0 <- p0[pzero < 0.5]
  uninf_covar <- uninf_covar[pzero < 0.5]

  # DE simulation
  if(nDE > 0){
    wts <- rowMeans(dat_alt)
    DE <- sample(1:nrow(dat_alt), nDE, prob = wts)
    truth[DE] <- TRUE

@@

  geneExp <- tbl_df(data.frame(pval=res_alt$pvalue, SE=res_alt$lfcSE, 
                               ind_covariate = res_alt$baseMean, 
                               effect_size=res_alt$log2FoldChange, 
                               test_statistic=res_alt$stat, qvalue=truth))

  if (uninformativeCovariate)
    geneExp <- mutate(geneExp, ind_covariate = uninf_covar)

@@

# Sim N5: Null Comparison 5v5

## Generate a list of SB results objects

## Plot average results over replications

# Sim N10: Null Comparison 10v10

## Generate a list of SB results objects

## Plot average results over replications

# Sim D5: DE Comparison 5v5

## Generate a list of SB results objects

## Plot average results over replications

res5d <- plotsim_standardize(de5, alpha = seq(0.01, 0.20, 0.01))

# Sim D10: DE Comparison 10v10

## Generate a list of SB results objects

## Plot average results over replications

res10d <- plotsim_standardize(de10, alpha = seq(0.01, 0.20, 0.01))

Here we compare the method ranks for the different sample sizes at alpha = 0.10.

```{r}
plotMethodRanks(c(resfile_d5, resfile_d10), 
                colLabels = c("DE 5", "DE 10"), 
                alpha = 0.10, xlab = "Comparison",
                excludeMethods = NULL)
```

# Uninformative Covariate

## Sim N5: Null Comparison 5v5

@@

if (!file.exists(resfile_n5_uninfCov)){
  null5 <- mclapply(X=1:B, rseed=rseed, FUN=simulateOneSplit, 
                    nDE=nDE, sampleSize=sampleSize, 
                   bd=bd, uninformativeCovariate = TRUE, mc.cores=nCores)
  saveRDS(null5, file=resfile_n5_uninfCov)
} else {
  null5 <- readRDS(file=resfile_n5_uninfCov)
}
```

@@

## Sim N10: Null Comparison 10v10

@@

if (!file.exists(resfile_n10_uninfCov)){
  null10 <- mclapply(X=1:B, rseed=rseed, FUN=simulateOneSplit, 
                     nDE=nDE, sampleSize=sampleSize, 
                   bd=bd, uninformativeCovariate = TRUE, mc.cores=nCores)
  saveRDS(null10, file=resfile_n10_uninfCov)
} else {
  null10 <- readRDS(file=resfile_n10_uninfCov)
}
```

@@

## Sim D5: DE Comparison 5v5

@@

@@

res5d <- plotsim_standardize(de5, alpha = seq(0.01, 0.20, 0.01))

@@

## Sim D10: DE Comparison 10v10

@@

res10d <- plotsim_standardize(de10, alpha = seq(0.01, 0.20, 0.01))

@@

## Sample size comparison

Here we compare the method ranks for the different sample sizes at alpha = 0.10.

```{r}
# uninformative covariate
plotMethodRanks(c(resfile_d5_uninfCov, resfile_d10_uninfCov), 
                colLabels = c("DE 5", "DE 10"), 
                alpha = 0.10, xlab = "Comparison",
                excludeMethods = NULL)
```

>>>>

# results files that will be generated below
# null
resfile_n5 <- file.path(resdir, "polyester-results-null5.rds")
resfile_n10 <- file.path(resdir, "polyester-results-null10.rds")

# strong covariate
resfile_d5 <- file.path(resdir, "polyester-results-de5.rds")
resfile_d10 <- file.path(resdir, "polyester-results-de10.rds")

# weak covariate 
resfile_d5_w <- file.path(resdir, "polyesterW-results-de5.rds")
resfile_d10_w <- file.path(resdir, "polyesterW-results-de10.rds")

# uninformative covariate
resfile_d5_uninfCov <- file.path(resdir, "polyester-results-de5-uninfCov.rds")
resfile_d10_uninfCov <- file.path(resdir, "polyester-results-de10-uninfCov.rds")

@@

```{r}
#' @param sampleSize is the number of samples in each condition
#' @param nDE is the number of DE genes
#' @param bd is the bench design object
#' @param uninformativeCovariate logical indicating whether to use an uninformative
#'  covariate (default is FALSE)
#' @param strongCovariate logical indicating whether to use a strongly informative
#'  covariate (default is TRUE)
#' @param rseed random seed
#' @param ngenes the number of genes to simulate. If this number is greater than 
#'  the number of genes in the count matrix, then the polyester parameters will be
#'  drawn with replacement from the available pool in order to obtain the requested
#'  number of genes.
simulateOneSplit <- function(X, rseed, 
                             nDE, sampleSize, bd, uninformativeCovariate = FALSE,
                             ngenes = nrow(counts),
                             strongCovariate = TRUE){

@@

  covar <- 1 / (1+exp(-runif(nrow(counts), 0, 10) + 5))

@@

  # filter genes with mostly zeroes
  pzero = rowSums(dat_alt==0)/ncol(dat_alt)
  dat_alt <- dat_alt[pzero < 0.5,]
  truth <- rep(FALSE, nrow(dat_alt))
  log2FC <- log2FC[pzero < 0.5]
  mu <- mu[pzero < 0.5]
  p0 <- p0[pzero < 0.5]
  covar <- covar[pzero < 0.5]

  # DE simulation
  if(nDE > 0){
    DE <- sample(1:nrow(dat_alt), nDE, prob = covar)
    truth[DE] <- TRUE

@@

  geneExp <- tbl_df(data.frame(pval=res_alt$pvalue, SE=res_alt$lfcSE, 
                               ind_covariate = covar, 
                               effect_size=res_alt$log2FoldChange, 
                               test_statistic=res_alt$stat, qvalue=truth))

  if (uninformativeCovariate){
    geneExp <- mutate(geneExp, ind_covariate = runif(length(covar)))
  }else if(!strongCovariate){
    geneExp <- mutate(geneExp, ind_covariate = pmin(1, abs(covar + rnorm(length(covar), 0, 0.25))))
  }

@@

# Null Comparisons

## N5: Null 5v5

@@

@@

## N10: Null 10v10

@@

@@

# Non-null Comparisons with Strongly Informative Covariate

## D5S: DE 5v5 Strong

@@

res5d <- plotsim_standardize(de5, alpha = seq(0.01, 0.10, 0.01))

## D10S: DE 10v10 Strong

res10d <- plotsim_standardize(de10, alpha = seq(0.01, 0.10, 0.01))

# Non-null Comparisons with Weakly Informative Covariate

## D5W: DE 5v5 Weak

Here we'll repeat the above, but using a weakly informative covariate.

if (!file.exists(resfile_d5_w)){
  de5 <- mclapply(X=1:B, rseed = rseed, FUN=simulateOneSplit, 
                  nDE=nDE, sampleSize=sampleSize, bd=bd, 
                  strongCovariate = FALSE, mc.cores=nCores)
  saveRDS(de5, file=resfile_d5_w)
} else {
  de5 <- readRDS(file=resfile_d5_w)
}
```

res5d <- plotsim_standardize(de5, alpha = seq(0.01, 0.10, 0.01))

## D10W: DE 10v10 Weak

Here we'll repeat the above, but for a DE comparison 10 versus 10 samples.

if (!file.exists(resfile_d10_w)){
  de10 <- mclapply(X=1:B, rseed = rseed, FUN=simulateOneSplit, 
                   nDE=nDE, sampleSize=sampleSize, bd=bd, 
                   strongCovariate = FALSE,
                   mc.cores = nCores)
  saveRDS(de10, file=resfile_d10_w)
} else {
  de10 <- readRDS(file=resfile_d10_w)
}

res10d <- plotsim_standardize(de10, alpha = seq(0.01, 0.10, 0.01))

# Non-null Comparisons with Uninformative Covariate

## D5U: DE 5v5 Uninformative

Here we'll repeat the above, but using a completely uninformative covariate.

@@

res5d <- plotsim_standardize(de5, alpha = seq(0.01, 0.10, 0.01))

@@

## D10U: DE 10v10 Uninformative

Here we'll repeat the above, but for a DE comparison 10 versus 10 samples.

@@

res10d <- plotsim_standardize(de10, alpha = seq(0.01, 0.10, 0.01))

@@

Here we compare the method ranks for the different sample sizes and informative
covariate settings at alpha = 0.10.

```{r}
plotMethodRanks(c(resfile_d5, resfile_d5_w, resfile_d5_uninfCov,
                  resfile_d10, resfile_d10_w, resfile_d10_uninfCov), 
                colLabels = c("DE5 S", "DE5 W", "DE5 U", "DE10 S", "DE10 W", "DE10 U"), 
                alpha = 0.10, xlab = "Comparison",
                excludeMethods = NULL)
```
